How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2020-12-07 analysis/paper/paper.Rmd
Not pretty but I could succeed to knit the whole paper!

>>>>

Our study focussed on two questions: how did the introduction of the new tools change foragers' mobility and landscape use; and how did the new technology change the way people use habitation sites? Overall, our results show that assemblages with stemmed points tend to have higher artefact densities and lower proportions of retouched pieces. Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.

## Artifact volumetric density and retouch frequency

We examined the artefact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns before and after the introduction of stemmed points. The analyses are based on the premise that highly mobile groups would carry lightweight toolkits with more retouched and easily replaceable composite tools, and leave a smaller amount of artefacts on site as a result of short-term occupations. These curated technologies invest time and effort in stone artefact manufacturing and maintenance to optimize travel costs for groups using high mobility or special purpose forays. On the other hand, less mobile groups would produce fewer retouched tools and more simple tools, and leave higher densities of artefacts generated from long-term occupations. These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes]. Our data support the concept of the curated and expedient technology by showing a strong pattern of artefacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies. Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artefact densities and fewer retouched pieces, compared to assemblages without stemmed points. In general, the new technology of stemmed points was more associated with expedient technological strategies.

It could be assumed that foragers with stemmed points were highly mobile groups and chose curated technologies since the stemmed points were likely used for hunting activities [@Chang_2013; @lee2019were]. However, our results show the opposite that the new technology was more often associated with  expedient technological strategies. Furthermore, our results show that different sites represent different occupation patterns during the time of technological transition, making it difficult to characterise this period with a single land use strategy. Seong claims that transitions from the expedient to curated technology in the Late Pleistocene were neither a straightforward nor unilineal processes and diverse site types can be more related to different occupation purposes including hunting camps, limited activity stations, caches, and so forth [@seong2015diversity]. Our findings support Seong, with some sites such as Bonggok and Yonghodong containing only 2 stemmed points, while Yongsandong has 38 points including broken tips and a base. More than 70% of Yongsandong stemmed points were unearthed, dominating the tool kit with 233 blades and other byproducts related to lithic manufacturing including cores and debris made of the same raw materials [@kim2004yongsan; @bae2012nature]. In the specific case of Yongsandong it seems likely that stemmed points were manufactured for hunting purposes, which dominated the function of the site. On the other hand, at Bonggok and Yonghodong stemmed points were only a minor part of activities there.

## Composition of toolkits and raw materials

The composition of toolkits during the Late Pleistocene of Korea has been described within multiple chronological sub-periods. Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artefacts; (2) mostly small quartzite and quartz vein artefacts with some large artefacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30ka [@seong2015diversity]. Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; ‘core tool industry with large flake knapping’ during early MIS 3 (59−40 ka), and ‘blade industry with tanged point chiefly’ during late MIS (40−24 ka) [@lee2012characteristics]. Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement. Our results, covering 49-24 ka and corresponding to the first three types of Seong’s divisions, support the continuation of core tools abundantly throughout the Late Pleistocene. In addition, our data show that side scrapers stayed as major tools while other tools occupied only a small portion of the assemblages. As stemmed points and blades increased in the toolkits, both choppers and side scrapers went through minor decreases but still remained abundant.

Changes in raw material composition were closely associated with the appearance of stemmed points. Previous studies claim that the existing tools, including choppers, polyhedrals, handaxes and cores were usually made of locally acquired quartzite, while the new tools including stemmed points and other blade assemblages were made of more fine-grained materials such as siliceous shale, hornfels, and obsidian which might be brought from distant sources [@Chang_2013; @seong2015diversity]. Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 41.5 ka, around the time that stemmed points appear. Projectile points, such as stemmed points, need high quality materials which enable high flaking quality, durability and effectiveness of an edge, as well as the creation of well-defined outline forms [@bamforth2009projectile]. The increased use of finer-grained raw materials likely changed forager mobility patterns to ensure their movements over the landscape supplied them with the materials needed to make stemmed points.

## Environmental and demographic context

We hypothesized that the introduction of stemmed points might reflect a preference for more portable and efficient hunting tools, as an adaptation to environmental or population changes. As the first composite tool appearing on the Korean Peninsula, stemmed points represent a major change in stone artefact technology. An important quality of composite tools is that the user can easily replace damaged parts,  contributing to an increased maintainability of the tool [@kuhn2015artifacts; @cardillo2010some]. Combining the stone projective with a wooden shaft further materials enhances functionality of the tool such as improving penetration by increasing weight [@browne1940projectile]. We expected that the functional advantage of the new composite tool might have allowed foragers to be more active in less productive landscapes. Previous studies simply describe the environment of the research period as  “overall cooler and drier during MIS 3” but lack detail due to a lack of direct paleoenvironmental proxies in Korea  [@Han_2008; @seong2008tanged; @Choi_2011; @Chang_2013; @im2015study; @bak2017late]. To explore environmental and demographic context related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene. The results show that groups with stemmed points occupied higher elevations, where the MAT was lower. This supports our hypothesis that stemmed points supported occupation of more marginal habitats. Through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures. We see a similar pattern of new technologies enabling expansion into marginal areas at a later time in north Paleo-Honshu Island, Japan, where the appearance of blade and projectile points at about 25 ka were associated with foragers moving into cold grassland areas [@Morisaki_Izuho_Terry_Sato_2015].

We investigated forager population dynamics to assess if population change might be a relevant mechanism to explain the appearance of stemmed points. The output of our SPD models indicate population increasing until 35 ka and remaining at the peak with minor deviations for about 3,000 years. If population dynamics were a key driver, then we expect population to be high at the time the new technology first appears. However, with the first stemmed points appearing at 38 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology. Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@Bae_2010]. Our results are consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points which may reflect a continuous influx of migrants over a long period.

# Conclusion

To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artefacts from 22 sites dated to 49-24 ka. The results show that the forager groups with stemmed points tended to be located at higher elevations, which means the use of stemmed points was associated with occupation of marginal or extreme environments. In addition, the groups with stemmed points had expedient technologies which reflect residential and less mobile land use patterns. The temperature gradually decreased during the technological transition and after the appearance of stemmed points, population gradually increased.

Our findings are limited by the relatively small number of sites and radiocarbon ages currently available. The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample. Furthermore, we currently lack functional data on individual artefact assemblages to support our conclusions about site use. With future work we may be better able to distinguish a high density of stone artefacts as results from a long-term/small group occupation versus a short-term/large group occupation. Further research should focus on detailed assemblage analysis at each site to better understand foragers mobility and site use strategies.
